---
output: pdf_document
---
# Statistical Inference Course Project: Part 1
By Neelajit Chandra


## Overview

The aim of this project is to investigate the exponential distribution in R and compare it with the Central Limit Theorem. The exponential distribution can be simulated in R with `rexp(n, lambda)` where
`lambda` is the rate parameter. The mean of exponential distribution is `1/lambda` and the standard deviation is also `1/lambda`. For this simulation, we set `lambda=0.2`. In this simulation, we investigate the distribution of averages of 40 numbers sampled from exponential distribution with `lambda=0.2`.


## Simulations

One thousand simulated averages of 40 exponentials have been calculated.

```{r}
set.seed(10)

lambda <- 0.2
num_sim <- 1:1000
sample_size <- 40

Means <- data.frame(x = sapply(num_sim, function(x) {
  mean(rexp(sample_size, lambda))
}))
```

The distribution of sample means is as follows.

```{r}
library(ggplot2)

# Theoretical Mean of distribution
th_mean <- 1/lambda
th_mean

# Sample Mean 
sim_mean <- mean(Means$x)
sim_mean

# Theoretical Variance of the distribution
th_var <- ((1/lambda)/sqrt(sample_size))^2
th_var

# Sample Variance
sim_var <- var(Means$x)
sim_var

# Plot of the histogram of means indicating the theoretical and sample values
ggplot(data = Means, aes(Means$x)) + geom_histogram(binwidth = 0.1) +
  geom_vline(mapping = aes(xintercept = c(1/lambda, mean(Means$x)),
                           linetype = c("Theoretical", "Simulated")),
             col = c("blue", "red"), show_guide = TRUE)

```

The distribution of sample means is centered at `r sim_mean`
and the theoretical center of the distribution is `r 1/lambda`.

The standard deviation of sample means is `r sd(Means$x)` whereas the expected standard deviation of the distribution is `r (1/lambda)/sqrt(40)`.

The variance of sample means is `r sim_var` whereas the theoretical variance of the distribution is `r ((1/lambda)/sqrt(40))^2`.


**Solutions**

1. The expected center (5.0) is very close to the distribution (`r mean(Means$x)`).  

2. The sample standard deviation (`r sd(Means$x)`) is also close to the expected standard deviation (0.7909).


## Distribution

From the plot of the distribution of sample means, we observe that it may be approximated as a normal distribution (with mean = 5 and standard deviation = 0.7909).

```{r}

ggplot(data = Means, aes(x = x)) +
  geom_histogram(aes(y = ..density..), fill = I("royalblue"),
                 binwidth = 0.2, color = I("black")) +
  stat_function(fun = dnorm, arg = list(mean = 5, sd = sd(Means$x)))

```
